Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-11 latex/main.tex
First pass at abstract

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{orcidlink}
\usepackage[super,comma,sort&compress]
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers

@@

\section*{SUMMARY}

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and
viral relationships can be represented by a single tree.
However, important SARS-CoV-2 recombinant lineages have emerged.
A more suitable approach is to use ancestral recombination graphs (ARGs),
which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
Recent computational advances have made ARG inference possible for datasets of millions of genome sequences.
Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and
reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences.
Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023.
We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution —
the backbone phylogeny, mutational spectra, and major recognized recombinants.
This ARG enables us to explore the evolutionary origins of known and novel recombinants.
Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and
can be analyzed efficiently and interactively using standard Python tools on a typical laptop.
This combination of capabilities makes sc2ts a powerful tool
for studying the recombination and mutation processes driving SARS-CoV-2 evolution.

@@

This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
but we found that it also excluded the major ``saltation'' events that characterize many VOCs.
These saltation lineages are characterized by a high number of accumulated mutations,
and are hypothesized to emerge from chronic infections
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}.
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron).
To distinguish time travelers with erroneous dates from true saltation lineages,
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
If on subsequent days these lineages form a sufficiently large group of samples,
sc2ts adds the group back to the ARG.
This ``retrospective sample grouping'' successfully captured the emergence of
most saltation lineages (Figure~\ref{fig:samples_per_day}).
The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
in which specific "seed" samples were identified for unconditional inclusion
(for details of how seed samples were chosen, see STAR Methods).
First arising in geographic regions with low sampling density,
the initial origins of these VOC are challenging to infer (REF).

@@

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions
in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
% TODO this is out of date and needs to be rewritten along with supplementary 
% analysis
Remarkably, given the differences in inference methods and data processing pipelines,
they are almost identically parsimonious,
with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ''X'', e.g., XA and XBB).
Below, we summarize our analysis of the Pango X lineages in the ARG
(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} and Document S2).

@@

@@

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.]

@@

@@

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an "origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e. the Pango Xs) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
% Referenced by the main Results.
The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
During site remapping, the 9-base deletion at 11283 was split into
a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).

<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).

\begin{table}
\centering
\caption{Major deletion events in the ARG.}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112410 & 44.8163 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340765 & 13.7286 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21988 & 8 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
28361 & 1 & N & 1246012 & BA.5.2.1 (Omicron) & 20822 & 0.8389 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Highly recurrent deletions.}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12713 \\
29762 & 1 & non-coding & 11356 \\
29779 & 1 & non-coding & 10925 \\
29769 & 1 & non-coding & 10783 \\
29774 & 1 & non-coding & 10707 \\
29700 & 1 & non-coding & 8543 \\
29614 & 1 & ORF10 & 5041 \\
29555 & 1 & non-coding & 4816 \\
29543 & 1 & non-coding & 4806 \\
29541 & 1 & non-coding & 4804 \\
203 & 1 & non-coding & 3602 \\
222 & 1 & non-coding & 3441 \\
22013 & 6 & S / S1-NTD & 2111 \\
21991 & 3 & S / S1-NTD & 2087 \\
28254 & 1 & ORF8 & 1967 \\
27556 & 64 & ORF7a & 1281 \\
27579 & 3 & ORF7a & 1014 \\
28093 & 1 & ORF8 & 898 \\
21765 & 6 & S / S1-NTD & 478 \\
27532 & 1 & ORF7a & 403 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Most commonly observed deletions in the ARG or Li et al. (2023).} 
\label{tab:common_dels}
\begin{tabular}{rrcrrrr}
\toprule
\multicolumn{3}{c}{} & \multicolumn{2}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
\cmidrule(lr){4-5} \cmidrule(lr){6-7}
Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%) \\
\midrule
28271 & 1 & non-coding & 1422469 & 57.3078 & 3853692 & 42.1183 \\
28248 & 6 & ORF8 & 1114712 & 44.9090 & 3026186 & 33.0742 \\
22029 & 6 & S / S1-NTD & 1114415 & 44.8970 & 3010665 & 32.9046 \\
11288 & 4 & ORF1ab / NSP6 & 871838 & 35.1242 & N/A & N/A \\
28362 & 9 & N & 871778 & 35.1218 & 3939164 & 43.0525 \\
21765 & 6 & S / S1-NTD & 825628 & 33.2625 & 3179887 & 34.7541 \\
21633 & 9 & S / S1-NTD & 532084 & 21.4364 & 1941962 & 21.2244 \\
11292 & 5 & ORF1ab / NSP6 & 531943 & 21.4307 & N/A & N/A \\
6513 & 3 & ORF1ab / NSP3 & 341714 & 13.7668 & 1984683 & 21.6913 \\
11283 & 5 & ORF1ab / NSP6 & 339935 & 13.6951 & 8360 & 0.0914 \\
21988 & 8 & S / S1-NTD & 339770 & 13.6885 & 590 & 0.0064 \\
22194 & 3 & S / S1-NTD & 313878 & 12.6454 & 1833050 & 20.0340 \\
11288 & 9 & ORF1ab / NSP6 & 311566 & 12.5522 & 3040576 & 33.2315 \\
21991 & 3 & S / S1-NTD & 308800 & 12.4408 & 912944 & 9.9779 \\
29762 & 1 & non-coding & 38030 & 1.5321 & N/A & N/A \\
29781 & 1 & non-coding & 36112 & 1.4549 & N/A & N/A \\
29769 & 1 & non-coding & 34315 & 1.3825 & N/A & N/A \\
29774 & 1 & non-coding & 33503 & 1.3498 & N/A & N/A \\
29779 & 1 & non-coding & 30515 & 1.2294 & N/A & N/A \\
11283 & 9 & ORF1ab / NSP6 & 60 & 0.0024 & 2011934 & 21.9891 \\
686 & 9 & ORF1ab / NSP1 & N/A & N/A & 115450 & 1.2618 \\
21987 & 9 & S / S1-NTD & N/A & N/A & 1918094 & 20.9635 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
% \includegraphics{sc2ts_methodology}
\caption{
Phylogenetic backbones: TODO
}
\label{fig:phylo_backbones}
\end{figure}

\begin{figure}
\centering
%TODO fix this to make the figures smaller.
\includegraphics[width=\linewidth]{mutational_spectra}
\caption{
All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
the mutation count data from Bloom et al. (2023)\cite{Bloom2023} (B), and
the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
}
\label{fig:mutational_spectra}
\end{figure}

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{orcidlink}
\usepackage[super,comma,sort&compress]
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers

@@

\section*{SUMMARY}
% The dataset is freakin amazing, and will be useful for all sorts of stuff
% for years to come
Millions of SARS-CoV-2 whole genome sequences were collected as part of the 
response to the COVID-19 pandemic, generating a dataset of unprecedented
depth, richness and potential for informing future studies.
% Trees are a useful abstraction of the data
Estimated genealogies, describing the likely inheritance structure among
the sampled sequences and their putative ancestors, are fundamental to
% JK: Not happy with this specific phrase, but it's good to emphasise the point
% that data is not insight, and the sheer volume of data here requires
% special effort.
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer such genealogies from viral 
genetic sequence data, however, is that recombination is negligible and 
the genealogy can be described as a tree.
% Recombination *is* important.
Recombinant lineages have risen to global prevalance in SARS-CoV-2,
and therefore such single-tree summaries of the data are necessarily
incomplete and potentially misleading.
% What do we do?
We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
that automaticaly detects and fully incorporates recombination, 
and apply it to 2.48 million genomes.
We show that this structure captures a broad range of evolutionary
features documented in the literature and provides novel 
insights into the relationships between previously identified
recombinant lineages.
% We need to emphasise the *resource*
Together, the inference method, inferred genealogy and 
a suite of efficient and flexible open-source analysis tools, 
represent a powerful resource upon which future research 
can build.

@@

This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
but we found that it also excluded the major ``saltation'' events that characterize many VOCs.
These saltation lineages are characterized by a high number of accumulated mutations,
and are hypothesized to emerge from chronic infections
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}.
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron).
To distinguish time travelers with erroneous dates from true saltation lineages,
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
If on subsequent days these lineages form a sufficiently large group of samples,
sc2ts adds the group back to the ARG.
This ``retrospective sample grouping'' successfully captured the emergence of
many saltation lineages such as Alpha (Figure~\ref{fig:subgraph_alpha}).
The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
in which specific ``seed'' samples were identified for unconditional inclusion
(for details of how seed samples were chosen, see STAR Methods).
First arising in geographic regions with low sampling density,
the initial origins of these VOC are challenging to infer (REF).
With the addition of these manual seeds, we successfully incorporated 
these challenging events (Figures~\ref{fig:subgraph_delta},
\ref{fig:subgraph_ba1_ba2}) and 
all other major lineages (Figure~\ref{fig:samples_per_day}) 
into the ARG.

@@

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled 
lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ''X'', e.g., XA and XBB).

@@

[NOTE: putting this here for now, until we find the right place to slot it
in next draft.]
See Document~\ref{sec:supplement_pango_lineages} for extended analyses,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
and Document S2 for ARG visualations of all Pango X lineages.

@@

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.
Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
labelled quadrants, and Figure~\ref{fig:copying_pattern_examples}
for examples of the corresponding copying patterns. Also reference
any additional examples that we include beyond A-D.]

@@

\subsubsection*{Comparison with UShER}
To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
properties but is more general (see the section ``Ancestral Recombination
Graphs``).
We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
from Figshare
and converted the ARG topology. We then downloaded the protobuf formatted
tree (tree.all\_viridian.202409.pb.gz), and exported the mutations to 
nucleotide format using \texttt{matUtils summary --translate}
(usher version 0.6.3), and translated
these mutations into tskit format. The resulting tszip file 
(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
mutations and requires 29MB of storage space.
We validated that the data was correctly encoded by comparing the 
variant data for the UShER tskit tree sequence against our alignments.
There was an exact match between the non-ambiguous nucleotide calls 
at 27469 sites. At the remaining 39 sites there were a significant 
number of gap characters (deletion calls), suggesting that the mismatches
could be due to differences in the alignments used as input to UShER.
See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
for details.

We then computed intersection ARGs for sc2ts and UShER by 
``simplifying''\cite{Kelleher2018efficient,Wong2024} 
both to the 2,475,418 samples and 27,507 sites that are shared.

All conversions were performed as part of the reproducible 
postprocessing pipeline described in [section?]. The full UShER
tree and the intersection files are available for download on Zenodo
in tszip format.

@@

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an ``origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e. the Pango Xs) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
By mapping deletions to the ARG, we identified
important deletion events (Table~\ref{tab:major_dels}),
many of which are associated with the origins of major VOCs,
for example, the deletions in ORF1ab and
S1-NTD (the N-terminal domain of Spike)
associated with the Alpha lineage.
We also identified highly recurrent deletions (Table~\ref{tab:rec_dels}),
for example, two known recurrent deletions in S1-NTD (REF).

Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.

\begin{table}
\centering
\caption{Major deletion events in the ARG. These events are inherited by at least 10,000 samples in the ARG. Count is the number of samples inheriting a deletion. The Pango lineage assigned to the node above which an event occurs is shown. Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Highly recurrent deletions. These are the top 20 deletions in terms of the number of occurrences in the ARG. Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Parsimony comparison of sc2ts and UShER}
\label{sec:supplement_sc2ts_usher_parsimony}

We performed an indepth analysis of the sc2ts and UShER ARGs 
(we refer to the UShER tree as an ARG for simplicity here as a single 
tree is also an ARG) in terms of parsimony.
A direct comparison is difficult because of the different aligments 
used as input, the different sets of sites used for inference and 
the fact that we are including deletions for a subset of sites in
the sc2ts ARG.

When we compare the intersection of the base sc2ts ARG (pre 
site-remapping), sc2ts is very slightly more parsimonious 
% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
with XX mutations over the YYY sites shared between them.
After site remapping, and including all mutations involving 
deletion characters, sc2ts has 143,690 mutations more 
(2,078,370 vs 1,934,680). If we exclude mutations involving 
deletions in sc2ts, this is reduced to 46,371 more (2.3\%)
than UShER.
If we restrict to the 26,192 sites that were not remapped, 
sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
and has around 2.5 times more mutations than UShER on the 
163 sites that were post-hoc remapped.

To provide a more objective measure of the overall parsimony 
of the inferred ARGs with respect to the Viridian alignments,
we remapped all 29903 sites in both ARGs 
using Fitch-Hartigan parsimony. Here the nucleotides
ACGT and ``-'' were treated as distinct characters, with all 
other values regarded as missing data. In this case, the sc2ts 
ARG has 629,864 (5.5\%) fewer mutations (10,840,687 
vs 11,470,551). Some of this difference may be due to the 
post-hoc incorporation of phylogenetic signal from deletions 
by sc2ts, but more work is required to fully understand 
the differences between the two approaches.

See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_usher\_sc2ts.ipynb}{analysis\_usher\_sc2ts.ipynb}
for details.

\subsection{Imputation comparison of sc2ts and UShER}
\label{sec:supplement_sc2ts_usher_imputation}
Both sc2ts and UShER support missing data in input alignments, and effectively
``hard call'' these values when the samples are integrated into the tree/ARG. 
In doing so, we impute the missing and ambiguous data for these samples.
We evaluate the imputation performance of sc2ts and UShER by comparing their
outputs on the Viridian alignments.

The total aligned dataset for 2,475,418 samples and 27,507 sites 
in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
Of the 162,838,461 missing data calls (Ns) in the alignments, sc2ts and UShER
disagreed in their imputed values for only 203,579 (0.125\%). 
Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed 435 (0.05\%) incorrectly (i.e., 
with a base that is not compatible with the
ambiguity code). 
UShER imputed only 7 calls (0.007\%) from this set incorrectly, likely 
due to special handling of ambiguous bases in the parsimony calculations.

This computation is a good example of type of integrated large-scale data analysis 
that can be easily performed using the combination of tskit and VCF Zarr.
The entire analysis is performed in a short Jupyter notebook, with the 
core calculation requiring about 30 lines of Python code. Retrieving 
data from the base alignments and ARGs for 2.48 million samples 
at the same time over all 27,507 sites took only 6 minutes, with a peak memory
usage of about 5GiB of RAM on a standard desktop computer.

See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/example\_data\_processing.ipynb}{example\_data\_processing.ipynb}
for details.

@@

\newcommand{\cprheight}{4.0em}

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
% FIXME: these samples are chosen randomly and so almost certainly do not
% belong to the correct quadrants.
A & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
B & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
C & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
D & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
\end{tabularx} 
\caption{Examples of problematic copying patterns. A-D are examples drawn from
the labelled quadrants in Figure~\ref{fig:recombinant_qc}.}
\label{fig:copying_pattern_examples}
\end{figure}

\renewcommand{\cprheight}{1.5em}
% Add some more space between rows. Note this it would be sligntly nicer if the 
% label was aligned vertically in the center of the row, but this is good
% enough for now.
\renewcommand{\arraystretch}{2}

\begin{figure}
\centering
\begin{tabularx}{\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
XC & \includegraphics[height=\cprheight]{static/XC.png}\\
XBR & \includegraphics[height=\cprheight]{static/XBR.png}\\
XA & \includegraphics[height=\cprheight]{static/XA.png}\\
XS & \includegraphics[height=\cprheight]{static/XS.png}\\
XL & \includegraphics[height=\cprheight]{static/XL.png}\\
XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
XF & \includegraphics[height=\cprheight]{static/XF.png}\\
XY & \includegraphics[height=\cprheight]{static/XY.png}\\
XG & \includegraphics[height=\cprheight]{static/XG.png}\\
XW & \includegraphics[height=\cprheight]{static/XW.png}\\
Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
\end{tabularx}
\caption{Copying patterns for the 20 recombination events associated with Pango 
X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order)}
\label{fig:pango_x_copying_patterns}
\end{figure}

% Reset this in case we make any more tables beyond this
\renewcommand{\arraystretch}{1}

\begin{figure}
\centering
\centering
\includegraphics[width=\linewidth]{tanglegram_base_tree.pdf}
\caption{
Tanglegram comparing phylogenetic backbones of sc2ts and UShER.
}
\label{fig:tanglegram_base}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{tanglegram_Delta_subtree.pdf}
\caption{
Tanglegram comparing sc2ts and UShER on the Delta subtree.
}
\label{fig:tanglegram_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{tanglegram_BA.2_subtree.pdf}
\caption{
Tanglegram comparing sc2ts and UShER on the BA.2 subtree.
}
\label{fig:tanglegram_ba2}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{tanglegram_BA.5_subtree.pdf}
\caption{
Tanglegram comparing sc2ts and UShER on the BA.5 subtree.
}
\label{fig:tanglegram_ba5}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}
\caption{
Subgraph of Alpha. [Some notes to help the reader know this 
reflects known events]
}
\label{fig:subgraph_alpha}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Delta.pdf}
\caption{
Subgraph of Delta. [Some notes to help the reader know this 
reflects known events]
}
\label{fig:subgraph_delta}
\end{figure}
